A Deterministic and Terminating Semantics for the Synchronous Programming Language~\CEU

### 2018-02-18 06-appendix.tex
Adiciona teorema sobre terminação NST ao apêndice

\begin{lemma}\label{lem.det-out}
  If~$\delta\out\delta_1$ and~$\delta\out\delta_2$ then~$\delta_1=\delta_2$.
\end{lemma}
\begin{proof}
  The lemma is vacuously true if~$\delta$ cannot be advanced by~$\out$
  transitions.  Suppose that is not the case and let~$\delta=\<p,n,e>$,
  $\delta_1=\<p_1,n_1,e_1>$ and~$\delta_2=\<p_2,n_2,e_2>$.  Then, there are
  two possibilities.
  \begin{case}
    $e\ne\nil$.  Both transitions are applications of rule~\R{push}.
    Hence~$p_1=p_2=\bcast(p,e)$, $n_1=n_2=n+1$, and~$e_1=e_2=\nil$.
  \end{case}
  \begin{case}
    $e=\nil$.  Both transitions are applications of rule~\R{pop}.
    Hence~$p_1=p_2=p$, $n_1=n_2=n-1$, and~$e_1=e_2=\nil$.\qedhere
  \end{case}
\end{proof}

@@

\begin{definition}\label{def.Hnst}
  A description~$\delta=\<p,n,e>$ is \emph{nested-irre\-ducible}
  iff~$p=\ceu{\Nop},\ceu{\Break}$ or~$\isblocked(p,n)$~is true.
  Nested-irreducible descriptions serve as normal forms for~$\nst$
  transitions: their program~$p$ and stack-level~$n$ embody the result of an
  exhaustive number of~$\nst$ transitions.  We will write~$\delta_\Hnst$ to
  indicate that description~$\delta$ is nested-irreducible.
\end{definition}

@@

\begin{lemma}\label{lem.irr-nst-n}
  If~$\delta\nst[n]\delta_\Hnst'$ then, for all~$i\ne{n}$, there is
  no~$\delta_\Hnst''$ such that~$\delta\nst[i]\delta''_\Hnst$.
\end{lemma}
\begin{proof}
  By contradiction.  Let~$\delta\nst[n]\delta'_\Hnst$, for some~$n\ge0$.
  There are two cases.
  \begin{case}\label{lem.irr-nst-n-case1}
    Suppose there are~$i>n$ and~$\delta''_\Hnst$ such
    that~$\delta\nst[i]\delta''$.
    %%
    Then, by definition of~$\nst[i]$,
    \begin{equation}\label{lem.irr-nst-n-eq1}
      \delta\nst[n]\delta'\nst[n+1]\delta_1'\nst[n+2]\cdots\nst[i]\delta''.
    \end{equation}
    Since~$\delta'=\<p',n,e'>$ is nested-irreducible,
    $p=\ceu{\Nop},\ceu{\Break}$ or~$\isblocked(p',n)$.  In either case, by
    the definition of~$\nst$, there is no~$\delta_1'$ such
    that~$\delta'\nst[1]\delta_1'$, which
    contradicts~\eqref{lem.irr-nst-n-eq1}.  Therefore, no such~$i$ can
    exist.
  \end{case}
  \begin{case}
    Suppose there are~$0\le{i}<n$ and~$\delta''_\Hnst$ such
    that~$\delta\nst[i]\delta''$.  Then since~$n>i$, by
    Case~\ref{lem.irr-nst-n-case1}, $\delta'$~could not exist, which is
    absurd.  Therefore, the assumption that there is such~$i$ is
    false.\qedhere
  \end{case}
\end{proof}

>>>>

\begin{lemma}\label{lem.det-out}
  If~$\delta\out\delta_1$ and~$\delta\out\delta_2$ then~$\delta_1=\delta_2$.
\end{lemma}
\begin{proof}
  The lemma is vacuously true if~$\delta$ cannot be advanced by~$\out$
  transitions.  Suppose that is not the case and let~$\delta=\<p,n,e>$,
  $\delta_1=\<p_1,n_1,e_1>$ and~$\delta_2=\<p_2,n_2,e_2>$.  Then, there are
  two possibilities.
  \begin{case}
    $e\ne\nil$.  Both transitions are applications of~\R{push}.
    Hence~$p_1=p_2=\bcast(p,e)$, $n_1=n_2=n+1$, and~$e_1=e_2=\nil$.
  \end{case}
  \begin{case}
    $e=\nil$.  Both transitions are applications of~\R{pop}.
    Hence~$p_1=p_2=p$, $n_1=n_2=n-1$, and~$e_1=e_2=\nil$.\qedhere
  \end{case}
\end{proof}

@@

\begin{definition}\label{def.Hnst}
  A description~$\delta=\<p,n,e>$ is \emph{nested-irre\-ducible}
  iff~$e\ne\nil$ or~$p=\ceu{\Nop},\ceu{\Break}$ or~$\isblocked(p,n)$~is
  true.  Nested-irreducible descriptions serve as normal forms for~$\nst$
  transitions: they embody the result of an exhaustive number of~$\nst$
  transitions.  We will write~$\delta_\Hnst$ to indicate that
  description~$\delta$ is nested-irreducible.
\end{definition}

@@

\begin{lemma}\label{lem.irr-nst-n}
  If~$\delta\nst[n]\delta_\Hnst'$ then, for all~$i\ne{n}$, there is
  no~$\delta_\Hnst''$ such that~$\delta\nst[i]\delta''_\Hnst$.
\end{lemma}
\begin{proof}
  By contradiction on the hypothesis that there is such~$i$.
  %%
  Let~$\delta\nst[n]\delta'_\Hnst$, for some~$n\ge0$.
  There are two cases.
  \begin{case}\label{lem.irr-nst-n-case1}
    Suppose there are~$i>n$ and~$\delta''_\Hnst$ such
    that~$\delta\nst[i]\delta''$.
    %%
    Then, by definition of~$\nst[i]$,
    \begin{equation}\label{lem.irr-nst-n-eq1}
      \delta\nst[n]\delta'\nst[n+1]\delta_1'\nst[n+2]\cdots\nst[i]\delta''.
    \end{equation}
    Since~$\delta'=\<p',n,e'>$ is nested-irreducible,
    $p=\ceu{\Nop},\ceu{\Break}$ or~$\isblocked(p',n)$.  In either case, by
    the definition of~$\nst$, there is no~$\delta_1'$ such
    that~$\delta'\nst[1]\delta_1'$, which
    contradicts~\eqref{lem.irr-nst-n-eq1}.  Therefore, no such~$i$ can
    exist.
  \end{case}
  \begin{case}
    Suppose there are~$0\le{i}<n$ and~$\delta''_\Hnst$ such
    that~$\delta\nst[i]\delta''$.  Then since~$n>i$, by
    Case~\ref{lem.irr-nst-n-case1}, $\delta'$~could not exist, which is
    absurd.  Therefore, the assumption that there is such~$i$ is
    false.\qedhere
  \end{case}
\end{proof}

The next lemma establishes some basic properties of sequences of~$\nst$
transitions.

\begin{lemma}\label{lem.props-nst-n}
  If~$\<p_1,n,e>\nst[n]\<p_1',n,e'>$ then, for any~$p_2$:
  \begin{enumerate}[(a)]
  \item\label{lem.props-nst-n.a}
    $\<\ceu{p_1;\,p_2},n,e>\nst[n]\<p_1';p_2,n,e'>$;
    %%
  \item\label{lem.props-nst-n.b}
    $\<\ceu{p_1\AtLoop{p_2}},n,e>\nst[n]\<\ceu{p_1'\AtLoop{p_2}},n,e'>$;
    %%
  \item\label{lem.props-net-n.c}
    $\<\ceu{{p_1}\AtAnd{p_2}},n,e>\nst[n]\<\ceu{{p_1'}\AtAnd{p_2}},n,e'>$;
    %%
  \item\label{lem.props-net-n.d}
    $\<\ceu{{p_1}\AtOr{p_2}},n,e>\nst[n]\<\ceu{{p_1}'\AtOr{p_2}},n,e'>$.
  \end{enumerate}
\end{lemma}
\begin{proof}
  By induction on~$n$.
  %%
  \begin{enumerate}[(a)]
  \item The lemma is trivially true for~$n=0$, as~$p_1=p_1'$, and follows
    directly from~\R{seq-adv} for~$n=1$.  Suppose
    \begin{equation}
      \label{lem.props-nst-n.a.eq1}
      \<p_1,n,e>\nst[1]\<p_1'',n,e''>\nst[n-1]\<p_1',n,e'>,
    \end{equation}
    for some~$n>1$.  Then~$\<p_1'',n,e''>$ is not nested-irreducible, i.e.,
    $e=\nil$ and~$p\ne{\ceu{\Nop},\ceu{\Break}}$ and~$\isblocked(p_1,n)$ is
    false.  By~\eqref{lem.props-nst-n.a.eq1} and by~\R{seq-adv},
    \begin{equation}
      \label{lem.props-nst-n.a.eq2}
      \<\ceu{p_1;\,p_2},n,e>\nst[1]\<\ceu{p_1'';\,p_2},n,e''>.
    \end{equation}
    From~\eqref{lem.props-nst-n.a.eq1}, by induction hypothesis,
    \begin{equation}
      \label{lem.props-nst-n.a.eq3}
      \<\ceu{p_1'';\,p_2},n,e''>\nst[n-1]\<\ceu{p_1';\,p_2},n,e'>.
    \end{equation}
    From~\eqref{lem.props-nst-n.a.eq2} and~\eqref{lem.props-nst-n.a.eq3},
    \[
      \<\ceu{p_1;\,p_2},n,e>\nst[n]\<\ceu{p_1';\,p_2},n,e'>.
    \]
  \end{enumerate}
\end{proof}

\begin{theorem}\label{thm.term-nst-*}
  For any~$\delta$ there is a~$\delta'_\Hnst$ such
  that~$\delta\nst[*]\delta'_\Hnst$.
\end{theorem}
\begin{proof}
  By induction on the structure of programs.
  %%
  Let~$\delta=\<p,n,\nil>$.  The theorem is trivially true if~$\delta$ is
  nested-irreducible itself, as by definition~$\delta\nst[0]\delta_\Hnst$.
  Suppose that is not the case.  Then, depending on the structure of~$p$,
  there are~11 possibilities.  In each one of them, we show that
  such~$\delta'_\Hnst$ indeed exists.
  \begin{case}
    $p=\ceu{\Mem(\Id)}$.
    %%
    Then, by~\R{mem},
    \[
      \<\ceu{\Mem(\Id)},n,\nil>\nst[1]\<\ceu{\Nop},n,\nil>_\Hnst\,.
    \]
  \end{case}

  \begin{case}
    $p=\ceu{\EmitInt(e)}$.
    %%
    Then, by~\R{emit-int},
    \[
      \<\ceu{\EmitInt(e)},n,\nil>\nst[1]\<\ceu{\CanRun(n)},n,e>_\Hnst\,.
    \]
  \end{case}

  \begin{case}
    $p=\ceu{\CanRun(n)}$.
    %%
    Then, by~\R{can-run},
    \[
      \<\ceu{\CanRun(n)},n,\nil>\nst[1]\<\ceu{\Nop},n,\nil>_\Hnst\,.
    \]
  \end{case}

  \begin{case}
    $p=\ceu{\Ifelse{\Mem(\Id)}{p'}{p''}}$.
    %%
    There are two subcases.
    \begin{subcase}
      $\eval(\ceu{\Mem(\Id)})$~is true.
      %%
      Then, by~\R{if-true} and by the induction hypothesis, there is
      a~$\delta'$ such that
      \begin{align*}
        \<\ceu{\Ifelse{\Mem(\Id)}{p'}{p''}},n,\nil>
        &\nst[1]\<p',n,e>\\
        &\nst[*]\delta'_\Hnst\,.
      \end{align*}
    \end{subcase}
    \begin{subcase}
      $\eval(\ceu{\Mem(\Id)})$~is false.
      %%
      Then, by~\R{if-false} and by the induction hypothesis, there is
      a~$\delta'$ such that
      \begin{align*}
        \<\ceu{\Ifelse{\Mem(\Id)}{p'}{p''}},n,\nil>
        &\nst[1]\<p'',n,e>\\
        &\nst[*]\delta'_\Hnst\,.
      \end{align*}
    \end{subcase}
  \end{case}

  \begin{case}
    $p=\ceu{p';\,p''}$.
    %%
    There are three subcases.
    \begin{subcase}
      \label{thm.term-nst-*.seq-nop}
      $p'=\ceu{\Nop}$.
      %%
      Then, by~\R{seq-nop} and by the induction hypothesis, there is
      a~$\delta'$ such that
      \[
        \<\ceu{\Nop;\,p''},n,\nil>
        \nst[1]\<p'',n,e>\nst[*]\delta'_\Hnst\,.
      \]
    \end{subcase}
    \begin{subcase}
      \label{thm.term-nst-*.seq-brk}
      $p'=\ceu{\Break}$.
      %%
      Then, by~\R{seq-brk} and by the induction hypothesis, there is
      a~$\delta'$ such that
      \[
        \<\ceu{\Break;\,p''},n,\nil>
        \nst[1]\<p'',n,e>
        \nst[*]\delta'_\Hnst\,.
      \]
    \end{subcase}
    \begin{subcase}
      $p'\ne\ceu{\Nop},\ceu{\Break}$.
      %%
      By induction hypothesis, there are~$p_1'$ and~$e$ such that
      \[
        \<p',n,\nil>\nst[*]\<p_1',n,e>_\Hnst\,.
      \]
      And by item~\eqref{lem.props-nst-n.a} of Lemma~\ref{lem.props-nst-n},
      \begin{equation}
        \label{thm.term-nst-*.seq-adv.eq1}
        \<\ceu{p';\,p''},n,\nil>\nst[*]\<\ceu{p_1';\,p''},n,e>.
      \end{equation}
      It remains to be shown that~$\<\ceu{p_1';\,p''},n,e>$ is indeed
      nested-irreducible.  There are four possibilities following from the
      fact that the simpler~$\<p_1',n,e>$ is nested-irreducible.
      %%
      \begin{subsubcase}
        $e\ne\nil$.  Then, by the definition of~$\Hnst$,
        description~$\<\ceu{p_1';\,p''},n,e>$ is also nested-irreducible.
      \end{subsubcase}
      \begin{subsubcase}
        $p_1'=\ceu{\Nop}$.
        %%
        From~\eqref{thm.term-nst-*.seq-adv.eq1},
        \[
          \<\ceu{p';\,p''},n,\nil>\nst[*]\<\ceu{\Nop;\,p''},n,e>\,.
        \]
        From this point on, this case is similar to
        Case~\ref{thm.term-nst-*.seq-nop}.
      \end{subsubcase}
      \begin{subsubcase}
        $p_1'=\ceu{\Break}$.
        %%
        From~\eqref{thm.term-nst-*.seq-adv.eq1},
        \[
          \<\ceu{p';\,p''},n,\nil>\nst[*]\<\ceu{\Break;\,p''},n,e>\,.
        \]
        From this point on, this case is similar to
        Case~\ref{thm.term-nst-*.seq-brk}.
      \end{subsubcase}
      \begin{subsubcase}
        $\isblocked(p_1',n)$ is true.
        %%
        Then, by definition,
        \[
          \isblocked(\ceu{p_1';p''},n)=\isblocked(p_1',n)=\mathit{true}\,.
        \]
        Thus, from~\eqref{thm.term-nst-*.seq-adv.eq1} and by the
        definition~$\Hnst$, $\<\ceu{p_1';\,p''},n,e>$~is also
        nested-irreducible.
      \end{subsubcase}
    \end{subcase}
  \end{case}

  \begin{case}
    $p=\ceu{\Loop{p'}}$.
    %%
  \end{case}

  \begin{case}
    $p=\ceu{p'\AtLoop{p''}}$.
  \end{case}

  \begin{case}
    $p=\ceu{{p'}\And{p''}}$.
  \end{case}

  \begin{case}
    $p=\ceu{{p'}\AtAnd{p''}}$.
  \end{case}

  \begin{case}
    $p=\ceu{{p'}\Or{p''}}$.
  \end{case}

  \begin{case}
    $p=\ceu{{p'}\AtOr{p''}}$.
  \end{case}
\end{proof}
